A thermodynamic view of dusty protoplanetary disks I: axisymmetric stability

### 2016-11-13 linear.tex
various minor edits

We solve the full dispersion relation, Eq. \ref{streaming_dispersion} for selected cases where SI  
growth rates have been 
calculated analytically from full two-fluid disk models and measured 
from direct numerical particle-gas simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_{x,z} = \eta r k_{x,z}$ where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also define the stopping time as
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

>>>>

We solve the full dispersion relation, Eq. \ref{streaming_dispersion} for selected cases where SI  
growth rates have been 
calculated analytically from full two-fluid disk models and confirmed
with particle-gas numerical simulations 
\citep[namely][]{youdin07b,bai10b}. These authors use normalized
wavenumbers $K_{x,z} = \eta r k_{x,z}$ where
\begin{align} 
  \eta \equiv -\frac{1}{2\rhog r\OmK^2}\frac{\p P}{\p r} = 
  \frac{1}{2\left(1-\tepsilon\right)}\frac{F_r}{r\OmK^2}, 
\end{align} 
measures the pressure offset of Keplerian rotation. We use
$\eta=0.05c_s/r\OmK$. They also define the stopping time as
$\tau_\mathrm{s}=\tstop/(1-\tepsilon)$.

### 2016-11-13 model.tex
various minor edits

%In the limit of strong dust-gas drag (stopping time $\tstop\to 0$) 
In the limit of strong dust-gas drag ($\tstop\to0$), \cite{laibe14}
give a single fluid description of the mixture, to
$O\left(\tstop\right)$, as:   
\begin{align} 
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H}_\mathrm{drag} + \mathcal{H}_\mathrm{eff}  - \Lambda, \label{tempeq} 
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}. 
\end{align}
For an ideal gas the pressure is given
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

@@

We now simplify the gas energy equation by considering short cooling 
times $\tcool\to 0$, appropriate for the outer parts of an irradiated
protoplanetary disk \citep{chiang97,lin15}. Then the disk temperature
$T = T_\mathrm{ref}$ at all times, and so we may
adopt a locally isothermal equation of state 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}(1 - \tepsilon)\rho,   
\end{align}
where $c_s(r,z)= \sqrt{\mathcal{R}T_\mathrm{ref}/\mu}$ is a prescribed
sound-speed profile fixed in time.

In practice we consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

If we define the effective sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
    \tepsilon\right)}$, then the equation of state has the standard
form $P=c_{s,\mathrm{eff}}^2\rho$. Since the dust-fraction $\tepsilon$
typically decrease away from the midplane, we expect vertically
isothermal dusty disks to in fact behave as if it
had a temperture \emph{increasing} away from $z=0$.

@@

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations. A variable $f$ is
subject to Eulerian perturbations of the form 
\begin{align}
 \real\left[ \delta f(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. For later convenience we
write 
\begin{align}
  \sigma = \ii s - \omega,
\end{align}
where $s$ and $\omega$ are the growth rate and real frequencies,
respectively. Thus perturbations have time dependence $e^{st +
  \ii\omega t}$.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. Then 
requiring $\sigma^2>0$  gives the criteria 
for the axisymmtric \emph{stability} of an isothermal gas perfectly 
coupled to dust,  
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. %We assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_p\ln{\left(1 -
    \tepsilon\right)}$ for the entropy in the standard
expression for said criteria \citep[e.g.]{tassoul78}.

@@

Eq. \ref{stability_est1}---\ref{stability_est2} are easily satisfied
in thin accretion disks where radial gradients are $O(1/r)$, 
$h_\mathrm{g}\ll 1$ and $\Htilde/H_\mathrm{g}$ not too
large.  Then the
magnitude of the second term on the left-hand-side  of either
inequality is much less than unity. Thus radially smooth, strictly
isothermal disks perfectly coupled with dust are stable against
axisymmetric perturbations.

@@

The physical interpretation of Eq. \ref{thermal_instability} is that 
work done by pressure forces in the dusty gas leads to growth in
oscillation amplitudes. To demonstrate this, we calculate the average
work done during a periodic oscillation, then show that if the work
done is positive, then the growth rate is positive, i.e. the
oscillation amplitude would actually grow.

@@

Noting that only products of perturbations contribute to
$\mathcal{W}$, we find that for periodic oscillations (time dependence 
$e^{\ii\omega t}$ and real $\omega$): 
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
where %the integral is taken over the volume of the
%mixture; and 
 the Lagragian perturbation $\Delta$ of a variable $f$ is 
$\Delta f = \delta f + \bm{\xi}\cdot\nabla f$; where $\bm{\xi}$ is the
Lagragian displacement ($    \xi_{x,z} =  \ii \dd v_{x,z}/\sigma$).  
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

Now, quite generally, Eq. \ref{lin_mass_full}---\ref{lin_energy_full} 
imply the integrand of the numerator in Eq. \ref{thermal_instability} 
is 
\begin{align} 
  \imag \left(\delta \mathcal{C}
  \nabla\cdot\dd \bm{v}^*\right) = 
  -|\sigma|^2\imag \left(\Delta P 
  \frac{\Delta\rho^*}{\rho}\right). \label{pdv}
\end{align}
%This states that `PdV' work (the right-hand-side) is 
%provided by dust-drag friction, and can be seen explicitly as 
%follows. 
Then combining Eq. \ref{work_real}, \ref{pdv} and
\ref{thermal_instability} 
gives  
\begin{align}
s = \frac{\mathcal{W}}{\mathcal{I}^2}, \label{growth_work}
\end{align}
where we have set $\sigma= - \omega$ in 
Eq. \ref{pdv}. Eq. \ref{growth_work} states that if the average work
done is positive during a periodic oscillation, $\mathcal{W}>0$, then
then its amplitude would actually grow ($s>0$).  %and not be periodic

Positive work done is done by a parcel if $-\omega \imag\left(\Delta  
P\Delta\rho^*\right)>0$. Without loss of generality, take $\omega>0$
and consider a mass element with $\Delta\rho = 1$. Then positive work 
requires $\imag\left(\Delta
P\right)<0$.  This corresponds to pressure perturbations \emph{lagging}
behind that in density (or displacements), since the phase of
the assumed perturbations increase in time.

@@

For dusty gas, the work done during oscillations 
 is attributed to finite dust-gas drag. The relative
drift between gas and dust causes a phase difference between the two
components.

@@

This thermodynamic interpretation does not explain 
\emph{why} drag forces causes gas pressure to lag behind the dust
density, but merely shows 
that this \emph{must} be the case for any growing modes associated the
dust-gas drag. To rigorously understand how dust-drag causes this 
lag requires an explicit solution to the linearized equations with 
detailed treatment of the function $\mathcal{C}$. However, given the complexity of
$\mathcal{C}$ (see Appendix \ref{lin_dust}), we might generally expect
the dusty disks to  support a range of stable and unstable modes, with the
latter being associated with pressure-density lag.

>>>>

In the limit of strong dust-gas drag, $\tstop\to0$, the mixture obeys 
the first-order (in $\tstop$) single-fluid equations: 
\begin{align} 
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}\\
  &\frac{D T}{D t} = - \left(\gamma-1\right)T\nabla\cdot\bm{v} +
  \mathcal{H}_\mathrm{drag} + \mathcal{H}_\mathrm{eff}  - \Lambda  \label{tempeq} 
\end{align}
\citep{laibe14}, where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative \emph{following the mixture} with center-of-mass velocity  
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}. 
\end{align}
For an ideal gas the pressure is given
by $P = \mathcal{R}\rhog T/\mu $, where $\mathcal{R}$ is
the gas constant and $\mu$ is the mean molecular weight.

@@

We now simplify the gas energy equation by considering short cooling 
times $\tcool\to 0$, appropriate for the outer parts of an irradiated
protoplanetary disk \citep{chiang97,lin15}. Then the disk temperature
$T = T_\mathrm{ref}$ at all times, and so we may
adopt a locally isothermal equation of state 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}\left(1 - \tepsilon\right)\rho,   
\end{align}
where $c_s(r,z)= \sqrt{\mathcal{R}T_\mathrm{ref}/\mu}$ is a prescribed
sound-speed profile fixed in time. In practice we consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

Notice Eq. \ref{eos} resembles an ideal gas equation of state but with  
effective temperature $T_\mathrm{eff} = 
T_\mathrm{ref}\left(1-\tepsilon\right)$. %sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
This effective mixture  
temperature is lowered with dust-loading. 
%   \tepsilon\right)}$. 
%then the equation of state has the standard
%form $P=c_{s,\mathrm{eff}}^2\rho$. 
Since $\tepsilon$ typically decrease away from the midplane, we expect
vertically isothermal dusty disks to in fact behave as if it had a
temperture \emph{increasing} away from $z=0$.

@@

In this section we discuss the basic properties of the 
dusty-gas mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations. A variable $f$ is
subject to Eulerian perturbations of the form 
\begin{align}
 \real\left[ \delta f(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. For later convenience we
write 
\begin{align}
  \sigma = \ii s - \omega,
\end{align}
where $s$ and $\omega$ are the growth rate and real frequencies,
respectively. Thus perturbations have time dependence $e^{st +
  \ii\omega t}$.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. Then 
requiring $\sigma^2>0$  gives the criteria 
for the axisymmtric \emph{stability} of an isothermal gas perfectly 
coupled to dust,  
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. %We assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_p\ln{\left(1 -
    \tepsilon\right)}$ for the entropy in the standard
expression for said criteria \citep{tassoul78}.

@@

Eq. \ref{stability_est1}---\ref{stability_est2} are easily satisfied
in thin accretion disks where radial gradients are $O(1/r)$, 
$h_\mathrm{g}\ll 1$ and $\Htilde/H_\mathrm{g}$ not too
large.  Then the
magnitude of the second term on the left-hand-side  of either
inequality is much less than unity. Thus radially smooth, strictly
isothermal disks perfectly coupled with dust are stable against
axisymmetric perturbations. This result holds for arbitrarily thin
dust layers.

@@

The physical interpretation of Eq. \ref{thermal_instability} is that 
work done by pressure forces in the dusty gas leads to growth in
oscillation amplitudes. To demonstrate this, we calculate the average
work done during a periodic oscillation, and show that if the work
done is positive, then the 
oscillation amplitude would actually grow.

@@

Noting that only products of perturbations contribute to
$\mathcal{W}$, we find that for periodic oscillations (time dependence 
$e^{\ii\omega t}$ and real $\omega$): 
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
where %the integral is taken over the volume of the
%mixture; and 
 the Lagragian perturbation $\Delta$ of a variable $f$ is 
$\Delta f = \delta f + \bm{\xi}\cdot\nabla f$ and $\bm{\xi}$ is the
Lagragian displacement ($    \xi_{x,z} =  \ii \dd v_{x,z}/\sigma$).  
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

Now, quite generally, Eq. \ref{lin_mass_full}---\ref{lin_energy_full} 
imply the integrand of the numerator in Eq. \ref{thermal_instability} 
is 
\begin{align} 
  \imag \left(\delta \mathcal{C}
  \nabla\cdot\dd \bm{v}^*\right) = 
  -|\sigma|^2\imag \left(\Delta P 
  \frac{\Delta\rho^*}{\rho}\right). \label{pdv}
\end{align}
%This states that `PdV' work (the right-hand-side) is 
%provided by dust-drag friction, and can be seen explicitly as 
%follows. 
Then combining Eq. \ref{work_real}, \ref{pdv} and
\ref{thermal_instability} 
gives  
\begin{align}
s = \frac{\mathcal{W}}{\mathcal{I}^2}, \label{growth_work}
\end{align}
where we have set $\sigma= - \omega$ in 
Eq. \ref{pdv}. Eq. \ref{growth_work} states that if the average work
done is positive during an oscillation, $\mathcal{W}>0$, then 
then its amplitude would actually grow ($s>0$).  %and not be periodic

Positive work is done by a fluid parcel if $-\omega \imag\left(\Delta  
P\Delta\rho^*\right)>0$. Without loss of generality, take $\omega>0$
and consider a mass element with $\Delta\rho = 1$. Then positive work 
requires $\imag\left(\Delta
P\right)<0$.  This corresponds to pressure perturbations \emph{lagging}
behind that in density (or displacements), since the phase of
the assumed perturbations increase in time.

@@

For dusty gas, the work done during oscillations 
 is attributed to finite dust-gas drag. The relative
drift between gas and dust causes a phase difference between the two
components, and hence between pressure and total density.

@@

This thermodynamic interpretation does not explain 
\emph{why} drag forces causes gas pressure to lag behind the dust
density, but shows 
that this \emph{must} be the case for any growing modes associated the
dust-gas drag. To rigorously understand how dust-drag causes this 
lag requires an explicit solution to the linearized equations with 
detailed treatment of the function $\mathcal{C}$. However, given the complexity of
$\mathcal{C}$ (see Appendix \ref{lin_dust}), we might generally expect
 dusty disks to  support a range of stable and unstable modes, with the
latter being associated with pressure-density lag.

### 2016-11-13 paper.tex
various minor edits

%\title{On the entropy of 
%  isothermal dusty gas: application to protoplanetary disks}  
\title{A thermodynamic view of dusty protoplanetary disks I:
axisymmetric stability}  
\author{Min-Kai Lin\altaffilmark{1} and Andrew N. Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to entropy conservation
  following an adiabatic gas.  
  Motivated by this similarity, we develop a thermodynamical
  framework to study dusty gas dynamics in protoplanetary disks. We  
 show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to show that 
\begin{inparaenum}[1)] 
\item 
 the settling of perfectly coupled dust cannot cause axisymmetric 
 instabilties, however thin the dust layer becomes; 
\item 
  the streaming instability neccessarily involves a pressure response
  that lags dust density; 
\item dust-loading generally stabilizes the vertical shear
  instability, and can be understood as a result of dust-induced
  buoyancy forces. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss future applications of the dusty/adiabatic gas
equivalence in finding dusty analogs of other hydrodynamic
instabilities, as well as a way to simulate dusty protoplanetary disks
using purely hydrodynamic codes with minimal effort.

@@

This paper is organized as follows. We first explain in  
\S\ref{grow_osc} how a pressure-density phase lag in fluids can lead to
instability. This motives us to find an analogy between dusty gas and
pure hydrodynamics, as it would allow us to understand dust dynamics
within the framework of standard hydrodynamics.

We begin our formal discussion in in \S\ref{setup} by listing the 
one-fluid equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} which are used in
explicit calculations. In \S\ref{limits} we discuss
general stability properties of dusty disks. We then move onto
explicit stablity analyses. In  
\S\ref{linear_problem} we list the linearized equations, and solve 
them for the SI in \S\ref{si}, and in 
 \S\ref{results} study dusty VSI. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

@@

From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure (i.e. restoring) forces is still 
increasing. This over-compensation causes the parcel to expand
beyond the maximum volume of the previous cycle. 
Similarly, from $C$ to $D$ the parcel is already contracting towards
equilibrium, but pressure 
forces are still dropping. This allows the  
parcel's contraction to over-shoot the maximum density attained in the
previous cycle. 
The overall positive work done leads to growth in the oscillation amplitude. 
This effect is similar to being pushed downwards on a swing on the way
down.

>>>>

%\title{On the entropy of 
%  isothermal dusty gas: application to protoplanetary disks}  
\title{A thermodynamic view of dusty protoplanetary disks I:
axisymmetric stability}  
\author{Min-Kai Lin\altaffilmark{1,2,3}}
\author{Andrew N. Youdin\altaffilmark{1}}
%\affil{Department of Astronomy and Steward Observatory, 
%University of
 % Arizona, Tucson, AZ 85721, USA\\
%Institute of Astronomy and Astrophysics, Academia Sinica, 
%Taipei 10617, Taiwan}
\altaffiltext{1}{Department of Astronomy and Steward Observatory, 
University of
Arizona, Tucson, AZ 85721, USA}
\altaffiltext{2}{Steward Theory Fellow}
\altaffiltext{3}{Institute of Astronomy and Astrophysics, Academia Sinica, 
Taipei 10617, Taiwan
}
\email{mklin@asiaa.sinica.edu.tw}

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to entropy conservation
  following an adiabatic gas.  
  Motivated by this similarity, we develop a thermodynamical
  framework to study dusty gas dynamics in protoplanetary disks. We  
 show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to show that 
\begin{inparaenum}[1)] 
\item 
 the settling of perfectly coupled dust cannot cause axisymmetric 
 instabilties, however thin the dust layer becomes; 
\item 
  the streaming instability neccessarily involves a pressure response
  that lags dust density; 
\item dust-loading generally stabilizes the vertical shear
  instability, which can be understood as a effect of dust-induced
  buoyancy forces. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss future applications of the dusty/adiabatic gas
equivalence in finding dusty analogs of other hydrodynamic
phenomena (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and to simulate dusty protoplanetary disks
using purely hydrodynamic codes.

@@

This paper is organized as follows. We first explain in  
\S\ref{grow_osc} how a pressure-density phase lag in fluids can lead to
instability. This motives us to find an analogy between dusty gas and
pure hydrodynamics, as it allows us to understand dusty-gas dynamics
within a thermodynamic framework.

We begin our formal discussion in \S\ref{setup} by listing the 
one-fluid equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} that are used in
explicit calculations. In \S\ref{limits} we discuss 
general stability properties of dusty disks. We then move onto
numerical stability analyses. In  
\S\ref{linear_problem} we list the linearized equations, and solve 
them for the SI in \S\ref{si}, and in 
 \S\ref{results} to study the dusty VSI. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

@@

From $A$ to $B$, a fluid parcel is expanding to return to 
equilbrium, but pressure (i.e. restoring) forces is still 
increasing. This over-compensation causes the parcel to expand
beyond the maximum volume of the previous cycle. 
Similarly, from $C$ to $D$ the parcel is already contracting towards
equilibrium, but pressure 
forces are still dropping. This allows the  
parcel's contraction to over-shoot the maximum density attained in the
previous cycle. 
The overall positive work done leads to growth in the oscillation amplitude. 
This effect is similar to being pushed downwards on a swing on
descent.

### 2016-11-13 summary.tex
various minor edits

The one-fluid formulation, 
Eqs. \ref{masseq}---\ref{tempeq}, can 
produce spuriously-growing modes with frequency $|\sigma|\gtrsim
\Omega$ (see Appendix \S\ref{spurious_epi}). This 
results from neglecting terms of $O(|\widetilde{\bm{v}}|^2)$ when converting
the full two-fluid equations to the one-fluid formulation. The full
one-fluid equations 
%is significantly more complex than our 
%model \cite[e.g.][]{laibe14}. It 
 includes an evolutionary equation for $\widetilde{\bm{v}}$, which has no
 hydrodynamic counter-part \citep{laibe14}. However, it is
 still possible to improve upon 
 the first-order formulation whilst retaining similarilty to  
 standard hydrodynamics.  
%without having the complexity of the full equations

@@

The conditions required for ConO might be realized in a locally
isothermal dusty disk. Since $\p_rP < 0$ in typical disk models, 
if the dust-induced radial entropy gradient is negative, then
$N_r^2  =  F_r \p_r\seff < 0$. In 
fact, typical disk models used for the streaming
instability have negligible background radial density gradient, then  
$N_r^2= -\rho F_r^2/P$.

@@

Since buoyancy is critical for the ZVI, it requires an almost
purely adiabatic gas ($\tcool\to\infty$). This limits its
applicability to the inner most parts of typical protoplanetary disks
\citep{lesur16}.

@@

%give numerical example 
We have applied our one-fluid stability analysis to find unstable modes  
similar to that described by \cite{loren15}. We use the same
parameters reported by them as follows. 
%we set $p=q=0$, which eliminates VSI and minimizes streaming
%instabilities as there is no radial pressure gradient at the disk
%midplane.   
The total metalicity is $Z=0.01$ and the dust scale-height $H_\mathrm{d} =
0.4H_\mathrm{g}$, giving a maximum dust-to-gas ratio
$\mathrm{max}(\epsilon) = 0.025$. The midplane stopping time is 
$t_\mathrm{s0}=0.02\OmK^{-1}$. The gas disk parameters are $(p,q,
\hgas) =(-1.5, -1,0.05)$.  We set $k = \pi/H_\mathrm{d}$.

>>>>

The one-fluid formulation, 
Eqs. \ref{masseq}---\ref{tempeq}, can 
produce spuriously-growing modes with frequency $|\sigma|\gtrsim
\Omega$ (see Appendix \S\ref{spurious_epi}). This 
results from neglecting terms of $O(|\widetilde{\bm{v}}|^2)$ after converting
the full two-fluid equations to the one-fluid formulation. The full
one-fluid equations 
%is significantly more complex than our 
%model \cite[e.g.][]{laibe14}. It 
 includes an evolutionary equation for $\widetilde{\bm{v}}$, which has no
 hydrodynamic counter-part \citep{laibe14}. However, it is
 still possible to improve upon 
 the first-order formulation whilst retaining similarilty to  
 standard hydrodynamics.  
%without having the complexity of the full equations

@@

The conditions required for ConO might be realized in a locally
isothermal dusty disk. Since $\p_rP < 0$ in typical disk models, 
if the dust-induced radial entropy gradient is negative, then
$N_r^2  =  F_r \p_r\seff < 0$. In 
fact, the typical disk models used to study the streaming
instability have zero background radial density gradient, in which
case  $N_r^2= -\rho F_r^2/P$.

@@

Since buoyancy is critical for the ZVI, it requires an almost
purely adiabatic gas ($\tcool\OmK\gg 1$). This limits its
applicability to the inner most parts of typical protoplanetary disks
\citep{lesur16}.

@@

%give numerical example 
We have applied our one-fluid stability analysis to find unstable modes  
similar to that described by \cite{loren15}. We use the same
parameters reported by them as follows. 
%we set $p=q=0$, which eliminates VSI and minimizes streaming
%instabilities as there is no radial pressure gradient at the disk
%midplane.   
The total metalicity is $Z=0.01$ and the dust scale-height $H_\mathrm{d} =
0.4H_\mathrm{g}$, giving a maximum dust-to-gas ratio
$\mathrm{max}(\epsilon) = 0.025$. The midplane stopping time is 
$t_\mathrm{s0}=0.02\OmK^{-1}$. The gas disk parameters are $(p,q,
\hgas) =(-1.5, -1,0.05)$.  We set $k_x = \pi/H_\mathrm{d}$.
